# (PART) Linear Systems {-}

# Linear Systems

A SMALL CHANGE. We have to load the package `pracma` to include the `rref` command.
```{r}
require(pracma)
```

To solve a linear system of equations $A x = b$, we first define the vector $b$ using the command below. The `c` command in R stands for **combine**. We are combining values into a vector.
```{r}
b = c(-2,10,4,-2)
b
```

Then we define the coefficient matrix $A$. Here we use `cbind` to combine the vectors into the columns of a matrix named $A$. You can use `rbind` if you want to combine the vectors into the rows of a matrix.
```{r}
A = cbind(c(1,2,-1,1),c(0,1,1,0),c(-1,2,1,-1),c(-1,1,0,-1),c(-1,5,0,-1))
A
```

We can augment the matrix A with b (we could have just made the full augmented matrix from the start, but this is a skill we will use later) by column binding A with b.
```{r}
Ab = cbind(A,b)
Ab
rref(Ab)
```
So the set of solutions in parametric form is

x = (1,2,3,0,0) + s (0,1,-1,1,0) + t (-1,1,-2,0,1)

and this is a "plane" in $\mathbb{R}^5$. It is in $\mathbb{R}^5$, because these vectors have 5 coordinates. It is a plane because it is spanned by two vectors that are not on the same line.  If you want to use $\LaTeX$ to typeset this answer, you do so as follows (see the .Rmd file for the code):
$$
\begin{bmatrix} x_1 \\ x_2 \\ x_3 \\ x_4 \\ x_5 \end{bmatrix}
= 
\begin{bmatrix} 1 \\ 2 \\ 3 \\ 0 \\ 0 \end{bmatrix}
+ s \begin{bmatrix} 0 \\ 1 \\ -1 \\ 1 \\ 0 \end{bmatrix}
+ t \begin{bmatrix} -1 \\ 1 \\ -2 \\ 0 \\ 1 \end{bmatrix}
$$

We can **verify** that our answer works by multiplying $A$ by one of the solutions above. Matrix multiplication uses the funny operation `%*%`.
```{r}
A %*% c(1,2,3,0,0)
```
Which equals $b$ just as we had hoped.

