

# Fractals 

```{r}
require(pracma)
```


You've probably heard of fractals: mathematical sets with a self-similar structure. Each smaller part is a miniature copy of the whole structure. Today, we will see how to use linear transformations and homogeneous coordiantes to generate a fractal.



We will generate a fractal by repeatedly  applying the following 3D linear transformations. 

```{r}
A1 = cbind(c(.86,-.03,0),c(.03,.86,0),c(0,1.5,1)) 
A2 = cbind(c(.2,.21,0),c(-.25,.23,0),c(0,1.5,1)) 
A3 = cbind(c(-.15,.25,0),c(.27,.27,0),c(0,.45,1)) 
A4 = cbind(c(0,0,0),c(.0,.17,0),c(0,0,1)) 
```


We pick one of these four transformations  using some randomness.  Most of the  time  we apply matrix  A1. Occasionally, we  apply matrix A2 or  A3. And rarely  we apply A4.


```{r}
get_lt <- function() {
  val = rand()
  if (val < .83) {
    A = A1
  } else if (val < .91) {
    A = A2
  } else if (val < .99) {
    A = A3
  } else {
    A = A4
  }
  
  A
}
```

So let's generate a fractal!

```{r,fig.height=5,fig.width=5}
pt = c(0,0,1)

plot(pt[1], pt[2], pch='.', xlim=c(-5,5), ylim=c(-1,12))
for (i in 1:1000) {
  points(pt[1], pt[2], pch='.')
  pt = get_lt() %*% pt
}

```


## Transforming My House.

So what is happening? Let's look at the effect of each of these linear transformations on my house

```{r,fig.height=5,fig.width=5}

house = cbind(c(0,0,1), c(0,3/4,1), c(1/2,3/4,1), c(1/2,0,1), c(1,0,1), c(1,1,1), c(5/4,1,1), c(0,2,1), c(-5/4,1,1), c(-1,1,1), c(-1,0,1), c(0,0,1));

(A = A2)
house2 = A^2 %*% house
(house2)

plot(house[1,],house[2,],type="n",xlim=c(-.125,.125),ylim=c(1,2.5),xlab="x",ylab="y")
abline(h=-6:6, v=-6:6, col="gray", lty="dotted")
#polygon(house[1,], house[2,], col = "gray", border = "blue")
polygon(house2[1,], house2[2,], col = "cyan", border = "black")
```

##  Watching a series of points.

Let's  see what happens when  we  repeatedly apply the same mapping.

```{r,fig.height=5,fig.width=5}
pt = c(0,0,1)

plot(pt[1], pt[2], pch=1, xlim=c(-2,2), ylim=c(-1,2))
for (i in 1:1000) {
  points(pt[1], pt[2], pch=1)
  pt = A1 %*% pt
}

```
##  Your Turn

Make some small adjustments to one of these matrices. Explore the impact of the fractal. Can you make it spikier?  Bushier?

```{r}
A1 = cbind(c(.86,-.03,0),c(.03,.86,0),c(0,1.5,1)) 
A2 = cbind(c(.2,.21,0),c(-.25,.23,0),c(0,1.5,1)) 
A3 = cbind(c(-.15,.25,0),c(.27,.27,0),c(0,.45,1)) 
A4 = cbind(c(0,0,0),c(.0,.17,0),c(0,0,1)) 
```

```{r,fig.height=5,fig.width=5}
pt = c(0,0,1)

plot(pt[1], pt[2], pch='.', xlim=c(-5,5), ylim=c(-1,12))
for (i in 1:1000) {
  points(pt[1], pt[2], pch='.')
  pt = get_lt() %*% pt
}

```


## Here are some from http://paulbourke.net/fractals/ifs/ 

             a     0.2020    0.1380
             b    -0.8050    0.6650
             c    -0.6890   -0.5020
             d    -0.3420   -0.2220
             e    -0.3730    0.6600
             f    -0.6530   -0.2770

```{r,fig.height=5,fig.width=5}

B1 = cbind(c(0.202,-.689,0),c(-.805,-0.342,0),c(-0.3730,-.653,1))
B2 = cbind(c(0.138,-.502,0),c(.665,-0.222,0),c(.66,-.2777,1))


get_bourke <- function() {
  val = rand()
  if (val < .6) {
    A = B1
  } else {
    A = B2
  }
  
  A
}


pt = c(0,0,1)

plot(pt[1], pt[2], pch='.', xlim=c(-.5,1), ylim=c(-1.25,.5))
for (i in 1:1000) {
  points(pt[1], pt[2], pch='.')
  pt = get_bourke() %*% pt
}
```

```{r}
get_list_ifs <- function(ifs) {
  vec <- list()
  for (i in 1:dim(ifs)[2]) {
    vec[[i]] = cbind(c(ifs[1,i],ifs[3,i],0),c(ifs[2,i],ifs[4,i],0),c(ifs[5,i],ifs[6,i],1))
  }
  vec
}

get_rand_idx <- function(p ) {
  ret_val  = 1
  q =  0
  x  = rand()
  for (i in 1:length(mat_list)) {
    q = q + p[i]
    if (x < q) {
      ret_val =  i
      break
    }
  }

  ret_val
}

bourke_dat = rbind(c(0.2020, 0.1380), c(-0.8050, 0.6650),
             c(-0.6890,-0.5020),
             c(-0.3420,-0.2220),
             c(-0.3730,0.6600),
             c(-0.6530,-0.2770))

bourke_list = get_list_ifs(bourke_dat)
bourke_p  =  c(.5,.5)


```

```{r}
dragon_list = rbind(c(0.824074,0.088272),
c( 0.281428,0.520988),
c(-0.212346,-0.463889),
c(0.864198,-0.377778),
c(-1.882290,0.785360),
c(-0.110607,8.095795))
dragon_p = c(0.8, 0.2)

dragon_list
```




```{r,fig.height=5,fig.width=5}
pt = c(0,0,1)
mat_list = get_list_ifs(dragon_list)
p = dragon_p

plot(pt[1], pt[2], pch='.', xlim=c(-7,7), ylim=c(0,12))

for (i in 1:1000) {
  points(pt[1], pt[2], pch='.')
  mat = mat_list[[get_rand_idx(p)]]
  pt = mat %*% pt
}
```
```{r,fig.height=3,fig.width=3}
spiral_data = rbind(c(0.787879,-0.121212, 0.181818),
c(-0.424242,0.257576,-0.136364),
c(0.242424,0.151515, 0.090909),
c(0.859848,0.053030, 0.181818),
c(1.758647,-6.721654, 6.086107),
c(1.408065,1.377236, 1.568035))



spiral_p = c(0.90,0.05, 0.05)

pt = c(0,0,1)
mat_list = get_list_ifs(spiral_data)
(mat_list)
p = spiral_p

plot(pt[1], pt[2], pch='.', xlim=c(-7,7), ylim=c(0,12))

for (i in 1:1000) {
  points(pt[1], pt[2], pch='.')
  mat = mat_list[[get_rand_idx(p)]]
  pt = mat %*% pt
}

```
