

# Dynamical Systems in 2D


Let $A$ be a square $n \times n$ matrix and let $\mathsf{x}_0 \in \mathbb{R}^n$. A **dynamical system** is a sequence of vectors $\mathsf{x}_0,\mathsf{x}_1,\mathsf{x}_2, \ldots, \mathsf{x}_t, \ldots$ where
$$
\mathsf{x}_{t} = A \mathsf{x}_{t-1} = A^t \mathsf{x}_0 \quad \mbox{for} \quad t \geq 1.
$$
The sequence $\mathsf{x}_0,\mathsf{x}_1,\mathsf{x}_2, \ldots, \mathsf{x}_t, \ldots$ is called the **trajectory** for initial vector $\mathsf{x}_0$.

A dynamical system evolves over time. The long-term behavior is governed by the eigenvalues of matrix $A$
We will look at some $2 \times 2$ dynamical systems to develop some intuition about eigensystems.

## Helper Function to Plot Dynamical Systems

I have written some code that makes some helpful plots. You should run this code chunk before the others.

```{r}
get_traj <- function(mat, x0, num) {
  traj = cbind(x0) 
  num
  for (i in 1:num) {
    traj = cbind(traj, mat %*% traj[,dim(traj)[2]])
    traj
  }
  
  return(traj)
}


plot_traj <- function(mat, x0, num) {
  traj = get_traj(mat,x0,num)
  points(traj[1,],traj[2,], pch=20, col=rainbow(length(traj)))
}

trajectory_plot <- function(mat, t=20, datamax=5, plotmax=10, numpoints=10, showEigenspaces=TRUE) {

  # initialize plot
  par(pty = "s")
  plot(c(0),c(0),type="n",
       xlim=c(-plotmax,plotmax),ylim=c(-plotmax,plotmax),
       xlab='x', ylab='y')
  abline(h=-plotmax:plotmax, v=-plotmax:plotmax, col="gray")
  
  mygrid <- expand.grid(x=seq(from = -datamax, by = 2*datamax/numpoints, l = numpoints+1),
                      y=seq(from =  -datamax, by = 2*datamax/numpoints, l = numpoints+1))


  for (t in 1:dim(mygrid)[1]) {
    plot_traj(A,c(mygrid[t,1],mygrid[t,2]),t)
  }
  
  if (showEigenspaces) {
    eigen = eigen(A)
    
    #mylabel = cat('lambda=', eigen$values[1], 'and lambda=', eigen$values[2])
    #title(xlab=mylabel)
    
    v1 = zapsmall(eigen$vectors[,1])
    v2 = zapsmall(eigen$vectors[,2])

    if (! class(v1[1]) == "complex") {
      if (v1[1] == 0) {
        abline(v=0)
      } else {
        abline(a=0,b=v1[2]/v1[1], col="blue")
      } 
      
      if (v2[1] == 0) {
        abline(v=0)
      } else {
        abline(a=0,b=v2[2]/v2[1], col="blue")
      }   
    }
    
  }
}
```



## Our first example

Let's start by looking at the example from the video
$$
A = \frac{1}{30}
\begin{bmatrix}
31 & 4 \\ 2 & 29
\end{bmatrix}.
$$

Here is some code that creates a single trajectory for $0 \leq t \leq 30$ starting at 
$\mathsf{x}_0 = 
\begin{bmatrix}
-0.5 & 1
\end{bmatrix}.$
The colors of the points follow the rainbow ordering at $t$ increases.


```{r, dynsys1, fig.height=4, fig.width=4, echo=TRUE}

A = 1/30 * cbind(c(31,2),c(4,29))
x0 = c(-0.5,1)

# initialize the plot
plot(c(0),c(0),type="n",xlim=c(-5,5),ylim=c(-5,5),xlab="x",ylab="y")
abline(h=-5:5, v=-5:5, col="gray")


plot_traj(A,x0,30)


```

We get a more complete picture when we plot multiple trajectories at once. 
So I have written a helper function to plot the trajectories of a grid of points. It also plots the eigenspaces for the matrix. You can specify 

* the matrix A
* the number of iterations
* the size of the square where the initial points lie
* the size of the plot
* the number of points along the side of the grid


```{r, dynsys2, echo=TRUE}
A = 1/30 * cbind(c(31,2),c(4,29))

trajectory_plot(A, t=30, datamax=5, plotmax=15, numpoints=5)

```

Perhaps this system is best understood by comparing what we see with the eigenvector and eigenvalues.



```{r, echo=TRUE}
eigen(A)
```

We can see that we have slight expansion along $[ 2, 1]^{\top}$ and slight contraction along $[-1,1]$. The long term behavior is an expansion in the direction of $[2, 1]^{\top}$.




## Your Turn

Now it's your turn to explore some dynamical systems. 

* Create trajectory plots for each of these dynamical systems.
* Characterize the long-term behavior. 
    + What direction to vectors converge to?
    + Do magnitudes increase? decrease? stabilize?
* Calculate the eigenvectors and eigenvalues and compare them to your plot.
    + The eigensystem should tell the same story as your plot.


If your original plot is confusing, **try changing the parameters** (initial square size, plot size, number of grid points). 

Here is some code for you to adapt for the examples.

```
'```{r, echo=TRUE}
A = cbind(c(1,0),c(0,1))

trajectory_plot(A, t=30, datamax=5, plotmax=15, numpoints=5)

eigen(A)

```


### Example 1

$$
A =  \frac{1}{20}
\begin{bmatrix}
24& -6 \\
1 & 19
\end{bmatrix}
$$

### Example 2

$$
A =  \frac{1}{60}
\begin{bmatrix}
55& -8 \\
-1 & 53
\end{bmatrix}
$$

### Example 3

$$
A =  \frac{1}{110}
\begin{bmatrix}
106& 12 \\
6 & 92
\end{bmatrix}
$$

### Example 4

$$
A =  \frac{1}{16}
\begin{bmatrix}
17& -15 \\
15 & 17
\end{bmatrix}
$$





