# Voting Patterns in the US Senate

In 2014, the [Pew Research Center published a report](https://www.pewresearch.org/politics/2014/06/12/political-polarization-in-the-american-public/) about the increasing polarization of US politics. They wrote:

> Republicans and Democrats are more divided along ideological lines – and partisan antipathy is deeper and more extensive – than at any point in the last two decades. These trends manifest themselves in myriad ways, both in politics and in everyday life.

Is this really true, or is it just hype? Let's see what linear algebra can tell us about the situation.



```{r}
df <- read.csv("/Users/abeverid/PycharmProjects/linearalgebra/data/Senate088matrix.csv", header=FALSE)
dim(df)
head(df)


sen <- read.csv("/Users/abeverid/PycharmProjects/linearalgebra/data/Senate088senators.csv", header=FALSE)

party = sen[,4]

color=rep("chartreuse", length(sen))
color[party=='D']="cornflowerblue"
color[party=='R']="firebrick"

mat = data.matrix(df)

sys = eigen(mat)


a1 = sys$values[1]
a2 = sys$values[2]
a3 = sys$values[3]

v1 = sys$vectors[,1]
v2 = sys$vectors[,2]
v3 = sys$vectors[,3]
v4 = sys$vectors[,4]


v1order= order(v2)
v2order= order(v2)


plot(v1, v2, col=color)
print (sys$values[1])
print (sys$values[2])

foo = data.frame(cbind(v1,v2, party))
rownames(foo)=sen[,2]

bar = foo[v1order,]
head(bar)

```

```{r}

sys$values

mat2 = mat - a1 * v1 %*% t(v1)

max(mat2)

heatmap(mat, Colv = NA, Rowv = NA, scale="none")
heatmap(mat2, Colv = NA, Rowv = NA, scale="none")
```

