
# Modeling of Ecological Systems

[Download this Rmd file](https://github.com/mathbeveridge/math236_f20/blob/main/09-ecological-models.Rmd)


```{r,message=FALSE}
# run this command to load some practical math functions
require(pracma)
```

## Northern Spotted Owl Population

We model the population dynamics of the Northern Spotted Owls using a Leslie Matrix. The age classes for the spotted owls are:

- juvenile (less than 1 year old)
- subadult (1 to 2 years old)
- adult (over 2 years old)

and the Leslie Matrix is:
$$L=\begin{bmatrix} 0&0&0.33 \\ 0.18 & 0 & 0 \\ 0 & 0.71 & 0.94 \end{bmatrix}.$$



### Population forecast

Suppose that we started out with 200 juveniles, 100 subadults and 500 adults. What happens to this population over time?
 
```{r, echo=TRUE}
L = cbind(c(0,.18,0),c(0,0,.71),c(.33,0,.94)) # the Leslie Matrix
L
start = c(200,100,500)  # the starting distribution
N = 50  # N is the number of iterations


```

This code shows the population dynamics. You shouldn't need to edit this.
```{r owls, fig.height=5, fig.width=7}
### this code follows the populations for N steps
m = dim(L)[1]  # m is the number of rows of L
X = matrix(0,nrow=m,ncol=N)  #  Store the results in a (m x N) matrix called X
X[,1] = start  # put start in the first column of X

# loop N times and put your results in X
for (i in 2:N) {
  X[,i] = L %*% X[,i-1]    # here is the loop
}

#### this code plots the populations in the (m x N) matrix X

t = seq(1,N)  # time

# Expand right side of clipping rect to make room for the legend
par(xpd=T, mar=par()$mar+c(0,0,0,7))

# Plot graph 
plot(t,X[1,],type='l',col=1,ylim=c(0,1000),ylab="population",xlab="time (year)", main="Population in Age Group")

for (i in 1:m) {
  lines(t,X[i,],col=i)
  points(t,X[i,],col=i,pch=20,cex=.8)
}

# Plot legend where you want
legend(60,1000,c("(0-1)", "(1-2)", "(over 2)"), col=1:m, lty = 1)

# Restore default clipping rect
par(mar=c(5, 4, 4, 2) + 0.1)
```
 
### The population at $t=100$

Change the code above to run for $N=100$ iteratons, and then $N=200$. Can you now make a stronger statement about the long-term population?



```{r owls2, fig.height=5, fig.width=7}

N = 200
### this code follows the populations for N steps
m = dim(L)[1]  # m is the number of rows of L
X = matrix(0,nrow=m,ncol=N)  #  Store the results in a (m x N) matrix called X
X[,1] = start  # put start in the first column of X

# loop N times and put your results in X
for (i in 2:N) {
  X[,i] = L %*% X[,i-1]    # here is the loop
}

#### this code plots the populations in the (m x N) matrix X

t = seq(1,N)  # time

# Expand right side of clipping rect to make room for the legend
par(xpd=T, mar=par()$mar+c(0,0,0,7))

# Plot graph 
plot(t,X[1,],type='l',col=1,ylim=c(0,1000),ylab="population",xlab="time (year)", main="Population in Age Group")

for (i in 1:m) {
  lines(t,X[i,],col=i)
  points(t,X[i,],col=i,pch=20,cex=.8)
}

# Plot legend where you want
legend(60,1000,c("(0-1)", "(1-2)", "(over 2)"), col=1:m, lty = 1)

# Restore default clipping rect
par(mar=c(5, 4, 4, 2) + 0.1)
```

Yes, we can now see that populations are dying out.

### Eigenvectors and eigenvalues of $L$

Use eigenvalues and eigenvectors to explain the asymptotic behavior of the system. You can use the function `eigen(A)` to compute the eigenvalues and eigenvectors of a matrix.


```{r, echo=TRUE}
# your code goes here
eigen(L)
```

The largest eigenvalue of less than one. This explains why the owls are dying: when the dominant eigenvalue is less than one, we tend to the zero vector as time increases.


## More recent numbers for Northern Spotted Owls

More recent spotted owl data give the following entries for the Leslie Matrix of the spotted owl population:

- Juvenile Survival 0.22
- Subadult Survival 0.85
- Adult Survival 0.94
- Subadult Fecundity 0.00
- Adult Fecundity 0.33


### Find the long-range populations

Using this new Leslie matrix $L$ and the same initial populations $\mathsf{x}_0 = [200,100, 500]$ as above, create a plot of the owl populations for $N=100$. 

```{r, echo=TRUE}
# Uncomment the next line and create this matrix 
# L = 

L = cbind(c(0,.22,0),c(.15,0,.85),c(.33,0,.94)) # the Leslie Matrix
L
start = c(200,100,500)
N = 100 
```


This code shows the population dynamics. You shouldn't need to edit this.

```{r owls3, fig.height=5, fig.width=7}
### this code follows the populations for N steps
m = dim(L)[1]  # m is the number of rows of L
X = matrix(0,nrow=m,ncol=N)  #  Store the results in a (m x N) matrix called X
X[,1] = start  # put start in the first column of X

# loop N times and put your results in X
for (i in 2:N) {
  X[,i] = L %*% X[,i-1]    # here is the loop
}

#### this code plots the populations in the (m x N) matrix X

t = seq(1,N)  # time

# Expand right side of clipping rect to make room for the legend
par(xpd=T, mar=par()$mar+c(0,0,0,7))

# Plot graph 
plot(t,X[1,],type='l',col=1,ylim=c(0,1000),ylab="population",xlab="time (year)", main="Population in Age Group")

for (i in 1:m) {
  lines(t,X[i,],col=i)
  points(t,X[i,],col=i,pch=20,cex=.8)
}

# Plot legend where you want
legend(60,1000,c("(0-1)", "(1-2)", "(over 2)"), col=1:m, lty = 1)

# Restore default clipping rect
par(mar=c(5, 4, 4, 2) + 0.1)
```

### The populations at $t=100$ versus the dominant eigenvector

 The following code compares your $t=100$ population ratios to the dominant eigenvector ratios. What do you notice? Why does this make sense?

```{r}

finalpop = X[,N]
finalpop = finalpop/sum(finalpop)

vec = as.numeric(eigen(L)$vectors[,1])
vec = vec/sum(vec)



tabledata = cbind(finalpop,vec)

tableframe = data.frame(tabledata)
names(tableframe) = c('population', 'eigenvector')
rownames(tableframe) = c('juveniles','subadults','adults')

knitr::kable(
  tableframe, booktabs = TRUE,
  caption = 'Ratios of vector entries'
)
```

These ratios are the same. This is as expected, since in the limit, the vectors should coverge to the direction of the dominant eigenvalue.


### The long-range prospects

What are the long-range prospects for the Northern Spotted Owl? Is it better or worse than the original data given above? 

They are much better. The population is increasing because the dominant eigenvalue 1.003 is larger than 1.

```{r}
eigen(L)$values[1]
```



## Brook Trout in Hunt Creek, MI

Here is the Leslie Matrix of a population of brook trout in Hunt Creek in Michigan. The population is categorized into 5 age categories: fingerlings (0,1), yearlings (1-2), young adults (2-3), adults (3-4), and adults (4-5). Right now the population is seen to be dying off. 

### The Leslie Matrix

Explain, in words, the meaning of each of the non-zero entries of the matrix $L$. 

```{r, echo=TRUE}
L = cbind(c(0,.05,0,0,0),c(0,0,.28,0,0),c(37,0,0,.16,0),c(64,0,0,0,.08),c(82,0,0,0,0.00))
L
```

### Find the long-range populations

Here are the population dynamics. Calculate the eigenvalues for $L$ and match what you see to the plot of he populations. What do you conclude about the long-range population?

```{r, echo=TRUE}
# your code goes here

eigen(L)

```

The largest eigenvalue 0.887 is less than 1. So the population is going to die out.
 
  
```{r trout, fig.height=5, fig.width=7, echo=TRUE}
start = c(100,100,100,100,100)  # the starting distribution
N = 30              # N is the number of iterations
m = dim(L)[1]
X = matrix(0,nrow=5,ncol=N)  #  Store the results in a 3 x N matrix called X
X[,1] = start  # put start in the first column of X

# loop N times and put your results in X
for (i in 2:N) {
  X[,i] = L %*% X[,i-1]    # here is the loop
}

t = seq(1,N)  # time

# Expand right side of clipping rect to make room for the legend
par(xpd=T, mar=par()$mar+c(0,0,0,7))

plot(t,X[1,],type='l',col=1,ylim=c(0,8000),ylab="population",xlab="time (year)",
     main="Population in Age Group")
for (i in 1:m) {
  lines(t,X[i,],col=i)
  points(t,X[i,],col=i,pch=20,cex=.8)
}
legend(25, 5000, legend=c("Fingerlings (0-1)", "Yearlings  (1-2)", "Young Adults (2-3)","Adults (3-4)","Adults (4-5)"), col=1:m, lty=1)


# Restore default clipping rect
par(mar=c(5, 4, 4, 2) + 0.1)

```

### River Clean Up

A river clean up effort is being conducted with the hope of *increasing the rate of survival from fingerlings to yearlings*. 

- To what level does that survival rate need to be increased to in order for the population to reach a steady state? Find your answer (accurate up to 3 decimal places) by trial and error. (You only need to change a single entry of $L$. Which one?)
- What are the population percentages in the stable population?

```{r, echo=TRUE}

# update the fingerling to yearling survival rate
L = cbind(c(0,.074,0,0,0),c(0,0,.28,0,0),c(37,0,0,.16,0),c(64,0,0,0,.08),c(82,0,0,0,0.00))
L

## do your eigenvalue/eigenvector analysis here
eigensys = eigen(L)
val = eigensys$values[1]
vec = eigensys$vectors[,1]

val
vec/sum(vec)

```

Changing the fingerling survival rate to 0.074 results in a dominant eigenvector of 1 (approximately). The long-range population percentages will match the dominant eigenvector calculated above:
$(0.9105, 0.0674, 0.0189, 0.0030, 0.0002)$.


This code shows the population dynamics. You shouldn't need to edit this. 


```{r trout2, fig.height=5, fig.width=7, echo=TRUE}
start = c(100,100,100,100,100)  # the starting distribution
N = 50              # N is the number of iterations
m = dim(L)[1]
X = matrix(0,nrow=5,ncol=N)  #  Store the results in a 3 x N matrix called X
X[,1] = start  # put start in the first column of X

# loop N times and put your results in X
for (i in 2:N) {
  X[,i] = L %*% X[,i-1]    # here is the loop
}

t = seq(1,N)  # time

# Expand right side of clipping rect to make room for the legend
par(xpd=T, mar=par()$mar+c(0,0,0,7))

plot(t,X[1,],type='l',col=1,ylim=c(0,8000),ylab="population",xlab="time (year)",
     main="Population in Age Group")
for (i in 1:m) {
  lines(t,X[i,],col=i)
  points(t,X[i,],col=i,pch=20,cex=.8)
}
legend(25, 5000, legend=c("Fingerlings (0-1)", "Yearlings  (1-2)", "Young Adults (2-3)","Adults (3-4)","Adults (4-5)"), col=1:m, lty=1)


# Restore default clipping rect
par(mar=c(5, 4, 4, 2) + 0.1)

```

