---
title: "Problem Set 4"
author: "Your Name Here"
date: "Due 11/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# 1. An Invertible Product of Rectangular Matrices

<!------------------------------------------------------>
a. Your answer 

<!------------------------------------------------------>
b. Your answer





# 2. Guessing the Inverse Matrix from a Pattern

<!------------------------------------------------------>
a. Your answer

<!------------------------------------------------------>
b. Your answer

<!------------------------------------------------------>
c. Your answer


# 3. LU Decomposition of an Invertible Matrix


<!------------------------------------------------------>
a. Your answer

<!------------------------------------------------------>
b. Your answer

<!------------------------------------------------------>
c. Your answer


<!------------------------------------------------------>
d. Your answer


```
B=cbind(c(1,2,3,4,5,6,7,8),
        c(0,2,3,4,5,6,7,8),
        c(0,0,3,4,5,6,7,8),
        c(0,0,0,4,5,6,7,8),
        c(0,0,0,0,5,6,7,8),
        c(0,0,0,0,0,6,7,8),
        c(0,0,0,0,0,0,7,8),
        c(0,0,0,0,0,0,0,8)
        )
```


# 4. Homogeneous Coordinates

Original picture

```{r,bird1,fig.height=5,fig.width=5, echo=TRUE}

bird = rbind(c(20, 21.3, 23.3, 21.6, 20.7, 21.7, 20.9, 20.3, 18.9, 18.1, 18.3, 19.6, 19.2, 19.1, 19.2, 19.4, 20), 
             c(20, 21.7, 20.9, 21.1, 19.3, 18.7, 18, 18.9, 18, 15.4, 18.6, 19.6, 20, 20.2, 20.3, 20.3, 20), 
             c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1))

fronds = rbind(c(10, 8, 6, 4, 2, 2.5, 4, 6, 9, 10, 10, 9, 7, 5, 2.5, 4, 6, 8, 9, 10, 10, 11, 11, 10.5, 12, 12.5, 12.5, 10, 10, 13, 15, 15.5, 15.5, 14, 12, 10, 10, 12, 14, 14, 13.5, 13, 12, 10),
               c(16, 16, 16, 15, 11, 15, 17, 17.5, 17, 16, 16, 17.5, 19, 19.5, 18.5, 20, 20.5, 20, 19, 16, 16, 18, 20, 20.5, 20, 19, 18, 16, 16, 17, 16, 15, 13.5, 15, 16, 16, 16, 15, 13, 11, 9, 11, 13, 16),
               c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1))

trunk = rbind(c(10, 9, 8, 7.5, 7.5, 8, 10, 8, 6.5, 6, 5, 5, 6, 7, 10),
              c(16, 14, 12, 10, 8, 5, 2.5, 2, 2, 3, 6, 9, 12, 14, 16),
              c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1))

# intialize the plot
plot(bird[1,],bird[2,],type="n",xlim=c(0,25),ylim=c(0,25),xlab="x",ylab="y")
abline(h=0:25, v=0:25, col="gray", lty="dotted")

# plot the bird, trunk and fronds
polygon(bird[1,], bird[2,], col = "gray", border = "blue")
polygon(fronds[1,], fronds[2,], col = "orange", border = "black")
polygon(trunk[1,], trunk[2,], col = "brown", border = "black")

```


Template for defining the linear transformations and creating an updated picture


```{r,bird2,fig.height=5,fig.width=5, echo=TRUE}

#####################
######## update this code with the appropriate linear transformations

birdmap = cbind(c(1,0,0),c(0,1,0),c(0,0,0))
leafmap = cbind(c(1,0,0),c(0,1,0),c(0,0,0))


#####################
######## you do not need to change this the code below this line

newbird = birdmap %*% bird
newfronds = leafmap %*% fronds


# initialize the plot
plot(newbird[1,],newbird[2,],type="n",xlim=c(0,25),ylim=c(0,25),xlab="x",ylab="y")
abline(h=0:25, v=0:25, col="gray", lty="dotted")

polygon(newbird[1,], newbird[2,], col = "cyan", border = "blue")
polygon(newfronds[1,], newfronds[2,], col = "green", border = "black")
polygon(trunk[1,], trunk[2,], col = "brown", border = "black")
```



Plots both pictures for comparison


```{r,bird3,fig.height=5,fig.width=5, echo=TRUE}
# initialize the plot
plot(newbird[1,],newbird[2,],type="n",xlim=c(0,25),ylim=c(0,25),xlab="x",ylab="y")
abline(h=0:25, v=0:25, col="gray", lty="dotted")

# plot the new bird and new fronds
polygon(newbird[1,], bird[2,], col = "gray", border = "blue")
polygon(newfronds[1,], fronds[2,], col = "orange", border = "black")

# plot the original bird, fronds  and trunk
polygon(newbird[1,], newbird[2,], col = "cyan", border = "blue")
polygon(newfronds[1,], newfronds[2,], col = "green", border = "black")
polygon(trunk[1,], trunk[2,], col = "brown", border = "black")

```
