---
title: "Template for Problem Set 7"
author: "Your Name Here"
date: "Due Date Here"
output:
  html_document: default
---


# Problem Set 7

**Note:** If you are using this PS 7 Template then you need to submit a PDF file.

* The easiest way is to (1) knit to HMTL, (2) click the "Open in Browser" button, (3) In the opened tab, "Print... Save as PDF"
 * There is a way to knit to PDF, but it is not working for me (error: missing framed.sty). I tried to install the missing resources, but I didn't get it to work. Given the effort involved, I think that the "easy way" HMTL -> Print to PDF is faster.

```{r, echo=FALSE}

# run this command to load some practical math functions
require(pracma)

# Creates a trajectory for the dynamical system
# A = the matrix
# start = the initial vector
# N = number of iterations
get_trajectory <- function(A, start, N) {
    ### this code follows the populations for N steps
  m = dim(A)[1]  # m is the number of rows of L
  X = matrix(0, nrow=m, ncol=N)  #  Store the results in a (m x N) matrix called X
  X[,1] = start  # put start in the first column of X
  
  # loop N times and put your results in X
  for (i in 2:N) {
    X[,i] = A %*% X[,i-1] 
  }
  
  return(X)
}


# Plots a trajectory along a time axis
# X  = the trajectory
# title = the title for the plot
# types = a vector of the names for each of the entries of the vector.
plot_trajectory <- function(X, title, types) {

  m = dim(X)[1]
  N = dim(X)[2]

  t = seq(1,N)  # time

  print(dim(X))
  print(dim(t))
    
  # Expand right side of clipping rect to make room for the legend
  par(xpd=T, mar=par()$mar+c(0,0,0,10))
  
  ymin = min(0, 1.1 * min(X))
  ymax = max(0, 1.1 * max(X))
  
  # Plot graph 
  plot(t, X[1,], type='l', col=1, ylim=c(ymin,ymax), ylab="amount", xlab="time",  main=title)
  
  for (i in 1:m) {
    lines(t, X[i,], col=i)
    points(t,X[i,], col=i, pch=20, cex=.8)
  }
  
  # Plot legend where you want
  legend(N *1.1, ymax * .85, types, col=1:m, lty = 1)
  
  # Restore default clipping rect
  par(mar=c(5, 4, 4, 2) + 0.1)
}

```


## Glucose-Insulin System Model


  a. My plot showing the trajectory and my observation about how long it takes for $\begin{bmatrix} G_{t} \\  H_{t} \end{bmatrix}$ to return  to homeostatis.  

  b. My answer about $\lambda = a \pm b i$ and $|\lambda| = \sqrt{a^2 +  b^2}$ and $\theta = \tan^{-1} (a/b)$. I'll include the R commands for these calculations by using an `{r echo=TRUE}' block like this one.
  
```{r, echo=TRUE}

# setting echo=TRUE means that the code will be shown in the knit output

a=1/2
b=sqrt(3)/2
atan(a/b) * 360 / (2 * pi)

```



## Population Interaction Models



a.  My answer about the constants $a,b,c,d$. I'll use the ``tabledata`` variable below.  I will replace each `'\t\t'` with the appropriate sentence **inside the quotemarks**, for example, `'a > 1'`


```{r, echo=FALSE}


tabledata = rbind(c('predation', '\t\t', '\t\t', '\t\t', '\t\t'),
                  c('competition', '\t\t', '\t\t', '\t\t', '\t\t'),
                  c('symbiosis', '\t\t', '\t\t', '\t\t', '\t\t'))


tableframe = data.frame(tabledata)
names(tableframe) = c('model', 'a', 'b', 'c', 'd')



knitr::kable(
  tableframe, booktabs = TRUE
)
```


b. My answer matching each of $A,B,C$ to one of the models. I could use a bullet list by editing the answer below.

* $A$ is predation/competition/symbiosis
* $B$ is predation/competition/symbiosis
* $C$ is predation/competition/symbiosis

c. My answer matching each of $A,B,C$ to one of the vector fields I could use a bullet list by editing the answer below.

* $A$ is Vector Field 1/2/3
* $B$ is Vector Field 1/2/3
* $C$ is Vector Field 1/2/3


d. My answer about the initial population is $[X_0, Y_0]^{\top} = [1, 2.5]^{\top}$ for each model. I could use a bullet list by editing the answer below.

- **Matrix $A$:** My description of the trajectory and its long term behavior, including the limiting population ratios $[a, b]^{\top}$.
- **Matrix $B$:** My description of the trajectory and its long term behavior, including the limiting population ratios $[a, b]^{\top}$.
- **Matrix $C$:** My description of the trajectory and its long term behavior, including the limiting population ratios $[a, b]^{\top}$.

##  Blue Whale Population Dynamics



```{r, echo=TRUE}
(L=rbind(c(0,0,.19,.44,.5,.5,.45),
         c(.77,0,0,0,0,0,0),
         c(0,.77,0,0,0,0,0),
         c(0,0,.77,0,0,0,0),
         c(0,0,0,.77,0,0,0),
         c(0,0,0,0,.77,0,0),
         c(0,0,0,0,0,.77,.78)))

```




a. My plot of the blue whales for 40 years, and my observation about whether this is population growth, equilibirum or  extinction,  


b. The R block with `{r echo=TRUE}` to calculate the eigensystem, and my conclusions about the growth rate and long term population ratios.


c. My final R block with `{r echo=TRUE}` that shows my value for $h$ and confirms that the populations are stable in the long run. My observation about the long-term population ratios.


## The Power Method for Eigenvalue Calculation

 

```{r, echo=TRUE}
A = cbind(c(4,2,4,8), c(2,2,4,4), c(4,4,9,1),c(8,4,1,5))

A

eigen(A)


estimate_dominant_eigenvector <- function(A, numiter) {
  n = dim(A)[1]
  x = rep(1,n)
  
  for (i in 1:numiter) {
    y = A %*% x
    maxval = max(y)
    minval = min(y)
    
    if (abs(maxval) > abs(minval)) {
      m = maxval
    }  else {
      m = minval
    }
    
    x = y/m
  }
  
  est = 'estimate'
  attr(est,'value') = m
  attr(est, 'vector') = x
  
  return(est)
}

estimate_nearest_eigenvector <- function(A, c, numiter) {
  n = dim(A)[1]
  
  B = A - c * diag(n)
  Binv = solve(B)

  est = estimate_dominant_eigenvector(Binv, numiter)
  
  val = attr(est, 'value')
  newval = 1/val+c
  
  attr(est,'value') = newval
  
  return (est)
}
```


a. My R code block with `{r echo=TRUE}` that shows the minimum `numiter` for `estimate_dominant_eigenvector' to converge to the right answer (up to 7 decimal places).


b. My answer to the four parts of this question.

    1. My answer that shows that  $(A - cI)v = (\lambda - c) v$

    2. My answer that shows that the zero vector is the only solution to $(A-cI)x=0$.

    3. My answer that shows $B v = \frac{1}{\lambda -c} v$.

    4. My answer that explains why $\frac{1}{\lambda -c}$ is the dominant eigenvalue.

&nbsp;

c. My R code block with `{r echo=TRUE}` which shows that `estimate_nearest_eigenvalue' does find the correct eigenvalue for an initial guess of $c$.


