---
title: "Template for Problem Set 8"
author: "Your Name Here"
date: "Due Date Here"
output:
  html_document: default
---


# Problem Set 8

**Note:** If you are using this template then you need to submit a PDF file.

* The easiest way is to (1) knit to HMTL, (2) click the "Open in Browser" button, (3) In the opened tab, "Print... Save as PDF"
 * There is a way to knit to PDF, but it is not working for me (error: missing framed.sty). I tried to install the missing resources, but I didn't get it to work. Given the effort involved, I think that the "easy way" HMTL -> Print to PDF is faster.

```{r, echo=FALSE}

# run this command to load some practical math functions
require(pracma)

```


## Two Orthogonality Properties


a. My proof for part (a) 

b. My proof for part (b)



## Bases for a Subspace and its Orthogonal Complement



a. My answer to part a. My code blocks use `{r echo=true}`.

b. My answer to part b. My code blocks use `{r echo=true}`.

c. My answer to part c. My code blocks use `{r echo=true}`.

d. My answer to part d. My code blocks use `{r echo=true}`.



## Eigensystem of a Symmetrix Matrix


a. My answer to part a. My code blocks use `{r echo=true}`.

b. My answer to part b. My code blocks use `{r echo=true}`.

c. My answer to part c. My code blocks use `{r echo=true}`.

d. My answer to part d. My code blocks use `{r echo=true}`.



## Modeling Fertility in 1888 Switzerland


```{r echo=TRUE}

fert = swiss[,1]
agric = swiss[,2]
exam = swiss[,3]
educ = swiss[,4]
cath = swiss[,5]
mort = swiss[,6]

```


a. My answer to part a. My code blocks use `{r echo=true}`.

b. My answer to part b. My code blocks use `{r echo=true}`.



## Cosine Similarity for US Senators

```{r, echo=TRUE}

library(readr)

senate.vote.file = "https://raw.github.com/mathbeveridge/math236_f20//main/data/SenateVoting109.csv"

record <- read_csv(senate.vote.file, col_names = TRUE)

clinton = record[record$Name == 'Clinton',]
collins = record[record$Name == 'Collins',]
frist = record[record$Name == 'Frist',]
mccain = record[record$Name == 'McCain',]
obama = record[record$Name == 'Obama',]
reid = record[record$Name == 'Reid',]


knitr::kable(
  head(record[,1:10],5), booktabs = TRUE,
  caption = 'First 5 rows of the 109th US Senate voting records'
)


```

a. My answer to part (a).

```{r, echo=TRUE}


# returns the vote vector for the senator
get_votes <- function(senator) {
  #unlist makes this dataframe row into a regular vector
  return (unlist(senator[,4:49])) 
}

# Implement this function!
# Return the magnitude of the senator's vote vector
get_vote_magnitude <- function(senator) {
  votes = get_votes(senator)
  
  # Your code goes here!
  magnitude = 1
  
  return (magnitude)
}

# Implement this function!
# Return the cosine of the angle between the vote vectors 
# of senator1 and senator2
get_vote_cosine_similarity <- function(senator1, senator2) {
  votes1 = get_votes(senator1)
  votes2 = get_votes(senator2)
  
  # your code goes here!
  cosine = 1
  
  return (cosine)
}

```

b. My answer to part b. My code blocks use `{r echo=true}`.


c. My answer to part c. 


```{r, echo=TRUE}



# Implement this function!
classify_senator <- function(senator) {

  # your code goes here!
  val = "R"
  
  return (val)
}

# returns a dataframe containing all misclassified senator names
# and their party affiliations
classify_all <- function() {

  mismatch = vector()
  
  for (i in 1:99) {
    senator = record[i,]
    party = classify_senator(senator)
    if (party !=  record[i,2]) {
      mismatch = c(mismatch, i)
    }
  }
  
  record[mismatch,1:2]
}


# Classify all senators!
classify_all()


```



